Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-04 agujournaltemplate.tex
Finalizing Discussion 85%

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months and $\geq$ 4 months), as these make the annual means less reliable. Particularly toward the later years of the time series, the sampling became sparse. The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see Figure \ref{fig:sup:correlation} in the appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibits marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period from 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval sees a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, which only recovers after 2011. The collapse in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is lost (see Figure \ref{fig:divts} $f$). 
Community evenness, as signified by the Pielou index, shows no seasonal signal and a very large variance within seasons and across depth (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data interpolated within the top 100 meters, we aimed to identify any apparent structure within the time series that could be related to the physical environment. To achieve this, we employed clustering analysis on both the annual aggregate of cell counts from microscopy and the individual monthly measurements. Figure \ref{fig:clustering} $a$ presents the results of clustering the annual aggregate communities using a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster encompassing the years 1996-2004 and 2014-2016, while the second cluster includes the intermediate years from 2004-2013. 
Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004. Turnover subsequently decreased during the period of cluster 2, but increased again in 2014, with return to cluster 1. This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.

@@

% Figure 6
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to determine if time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased toward larger cell sizes, and therefore, most of the predicted genus cell counts are from diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black line indicates cumulative importance across the entire community, and the colored dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range where the community changed most, indicating these values as a threshold for community composition. See figure \ref{fig:sup:GFoutput_lags_extra} in the appendix for response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W}
\label{fig:map}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
\caption{Heatmap of yearly average values, normalized for comparison as z-scores. Red indicates a larger value, blue indicates a smaller value for that particular year. The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with two or less and three or more missing values. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\label{fig:zscore}
\end{figure}

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals. (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season. Upwelling season lasts from December to May, Rainy Season from June to November.}
\label{fig:divts}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
\caption{(a) Hierarchichal lustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.}
\label{fig:clustering}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure5_ClustCompPlot_v2.pdf}
\caption{(Plots (a-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (p-value: ***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
\label{fig:clustcomp}
\end{figure}

@@

\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters. Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & **\\
Wind speed & 7074.0 & 0.143 & n.s.\\
Precipitation & 6251.0 & 0.004 & **\\
Evaporation & 9979.0 & 3.68e-5 & ***\\
\addlinespace
21 °C Isotherm & 5293.0 & 0.034 & *\\
SST & 4712.0 & 0.001 & ***\\
Salinity & 7636.0 & 0.002 & **\\
$NO_3$ & 6378.0 & 0.018 & *\\
$PO_4$ & 7359.0 & 0.012 & *\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
\addlinespace
Chlorophyll $a$ & 8229.0 & 9.63e-6 & ***\\
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & *\\
\bottomrule
\end{tabular}
\end{table}

@@

This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}). Concurrently, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine captures observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.

When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. \citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories. Here, we focus on the more consistently sampled fluorometric measurements. Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.

Despite some difficulties in reconciling HPLC pigment data and microscopy cell counts, both methods indicate a reduction in diatom abundance after 2004. Cell counts of other functional groups decreased, while HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco Basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.

% check sizes of cells, then work this into a paragraph with another sentence!!
If we compare the two parts of the first cluster, the community reorganized, particularly as diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.

However, the shift in yearly mean genus richness does not correspond with the abundances in the cell counts. Instead, we observe a marked reduction following 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014. The same trend is evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. %

The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community. This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}. We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly, an increase a few years prior to the return in abundance to previous levels. 
This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
We observe a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
To identify the variables that can best predict changes in the phytoplankton community, we conducted a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices, along with in situ measurements of bottom-up drivers, as predictors. Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series. Interestingly, our analysis revealed that the strongest predictors of changes within the community are not in situ measurements, but rather the AMO and the MEI v.2 climate indices. These are followed in importance by nitrate concentration and SST (see Figure \ref{fig:GF}).

>>>>

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months and $\geq$ 4 months), as these make the annual means less reliable. Particularly toward the later years of the time series, the sampling became sparse. The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in the appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibits marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period from 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval sees a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, which only recovers after 2011. The collapse in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is lost (see Figure \ref{fig:divts} $f$). 
Community evenness, as signified by the Pielou index, shows no seasonal signal and a very large variance within seasons and across depth (see Figure \ref{fig:divts} $h$). The inter-annual dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data interpolated within the top 100 meters, we aimed to identify any apparent structure within the time series that could be related to the physical environment. To achieve this, we employed clustering analysis on both the annual aggregate of cell counts from microscopy and the individual monthly measurements. Figure \ref{fig:clustering} $a$ presents the results of clustering the annual aggregate communities using a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster encompassing the years 1996-2004 and 2014-2016, while the second cluster includes the intermediate years from 2004-2013. 
Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004. Turnover subsequently decreased during the period of cluster 2 but increased again in 2014, marking the return to cluster 1. This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster respectively. The diatom Thalassiosira dominated the early period and was abundant in all other periods. Pseudo-nitzschia dominates cluster 2, where the haptophyte Emiliana was the second most abundant genus. The later part of cluster 1 is dominated by Emiliana.

@@

% Figure 6
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to determine if time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased toward larger cell sizes, and therefore, most of the predicted genus cell counts are from diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black line indicates cumulative importance across the entire community, and the colored dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range where the community changed most, indicating these values as a threshold for community composition. See Figure \ref{fig:sup:GFoutput_lags_extra} in the appendix for response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Map of CARIACO Ocean Time-Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W.}
\label{fig:map}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
\caption{Time series showing annual z-scores of selected indicator and driver variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies for that year. Deeper shades indicate larger anomalies. All data was available at monthly resolution. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to lack of data, additionally we marked years with more than two and four or more missing monthly values to highlight unreliable annual mean estimates. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\label{fig:zscore}
\end{figure}

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals. (a,c,e,g) Large dots and lines mark annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season. Upwelling season lasts from December to May, Rainy Season from June to November.}
\label{fig:divts}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
\caption{(a) Hierarchichal lustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows how environmental variables are co-varying with the community data. Length of vector indicates strength of relationship.}
\label{fig:clustering}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure5_ClustCompPlot_v2.pdf}
\caption{(Plots (a-o) below show density distributions of the environmental variables, chlorophyll \textit{a} and diversity indices between the clustered years. Dashed line marks the median. Significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (p-value: ***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
\label{fig:clustcomp}
\end{figure}

@@

\caption{Results of two-sided Wilcoxon sum rank test with continuity correction for variables between the two clusters. Cluster 1 is defined as the years 1996 - 2003 and 2014 - 2016. Cluster 2 includes the years 2004 - 2013.}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & **\\
Wind speed & 7074.0 & 0.143 & n.s.\\
Precipitation & 6251.0 & 0.004 & **\\
Evaporation & 9979.0 & 3.68e-5 & ***\\
\addlinespace
21 °C Isotherm & 5293.0 & 0.034 & *\\
SST & 4712.0 & 0.001 & ***\\
Salinity & 7636.0 & 0.002 & **\\
$NO_3$ & 6378.0 & 0.018 & *\\
$PO_4$ & 7359.0 & 0.012 & *\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
\addlinespace
Chlorophyll $a$ & 8229.0 & 9.63e-6 & ***\\
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & *\\
\bottomrule
\end{tabular}
\end{table}

@@

% Clearly and briefly restating main findings 
% Interpret what these findings mean in the context of the study
% Emphasize what is novel and important (without comparison)
% Only in following paragraphs: Compare to existing literature, highlight agreements, contradictions, and new persepctives. Discuss why and explain reasons for either.

\subsection{Community reorganization and biomass shifts}
% P1: Introductory paragraph highlighting the overall dynamics/shifts
This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is in turn influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}). Concurrently, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine captures observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.

% P2: Discuss community reorganisation, biomass shifts vs HPLC data
We found that the annual aggregates of community data showed two clusters, one including the time series up until 2004 and the end of the time series past 2013, and the other consisting of the intermediate time period between 2004 and 2013 (see Figure \ref{fig:clustering}).  
When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}, as was observed in the Cariaco Basin.

% Disucssion of HPLC vs Microscopy!
When comparing microscopy counts to previous analyses of HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}. Cell counts of all functional groups decreased, while HPLC data indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. \citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories. 
Here, we focus on the more consistently sampled fluorometric measurements. Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. Changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.

% P3: Shift in Diversity does not match shift in biomass/cell counts, why?
However, the shift in yearly mean genus richness does not correspond with the abundances in the cell counts. Instead, we observe a marked reduction from 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014. The same trend is evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community. This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}. We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly, an increase a few years prior to the return in abundance to previous levels. 
This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
We observe a significant difference in the Pielou's index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pielou's index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community. The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are primarily driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
